Peer Assessment-1 (Activity Monitoring)
==============================
##Loading and pre-processing the Activity Monitoring Data. 
Following steps are done in preprocessing

- Unzipped the file using unzip command

- Loaded CSV file "activity.csv" using read.csv command into data frame Data.Activity
```{r}
rm(list = ls())
DataDir="/Users/Rahela/Documents/Data Science/R Programs/Reproducible research/RepData_PeerAssessment1"
setwd(DataDir)
if(!file.exists("activity.csv")){unzip("activity.zip")}
Data.Activity=read.csv("activity.csv")
Data.Activity$interval=as.factor(Data.Activity$interval)
summary(Data.Activity)
```
It is observed that there are three variables in the data i.e., steps, date and interval. Also there are "NA"s in the variable "steps".


```{r,echo=TRUE}
Data.Activity.NoNA=Data.Activity[complete.cases(Data.Activity$steps),]
library(plyr)
Steps.Day.Summary=ddply(Data.Activity.NoNA,"date",summarise,Total.Steps.InDay=sum(steps))
#print(Steps.Day.Summary)
A=as.table(summary(Steps.Day.Summary$Total.Steps))
binwidth=(A[6]-A[1])/20#(Max-Min)/200
```
## To plot histogram and reporting mean and median
```{r, echo=TRUE}
library(ggplot2)
qplot(Steps.Day.Summary$Total.Steps.InDay,binwidth=binwidth,geom = "histogram",main="Histogram of Steps/Day",xlab = "Steps",ylab = "Count (Days)",fill=I("blue"), col=I("red"))
Mean.Steps.InDay=mean(Steps.Day.Summary$Total.Steps.InDay)
Median.Steps.InDay=median(Steps.Day.Summary$Total.Steps.InDay)
```
Average (mean) number of steps per day is `r Mean.Steps.InDay` and Median steps per day is `r Median.Steps.InDay`

## Average daily activity pattern
```{r, echo=TRUE}
Steps.Minute.Summary=ddply(Data.Activity.NoNA,"interval",summarise,Average.Steps=mean(steps))
Steps.Minute.Summary$interval=as.integer(as.character(Steps.Minute.Summary$interval))
Max.Int=Steps.Minute.Summary$interval[which.max(Steps.Minute.Summary$Average.Steps)]
qplot(interval,Average.Steps,data = Steps.Minute.Summary,geom = "line",main = "Average steps in an interval",xlab = "Interval",ylab = "Average Steps")
```

The interval in which we see maximum activity is `r Max.Int`.
## Dealing with Missing values
```{r}
Total.NAs=sum(is.na(Data.Activity$steps))
```

Total "NA"s in the data of steps are `r Total.NAs`. In the following the missing values i.e. NAs are replaced with the average number of steps for the similar interval in the entire data.

```{r}
Data.Activity.Mod=Data.Activity
Data.Activity.Mod$Steps.Interval.Avg=Steps.Minute.Summary$Average.Steps[match(Data.Activity.Mod$interval,Steps.Minute.Summary$interval)]
Data.Activity.Mod$steps=ifelse(is.na(Data.Activity.Mod$steps),Data.Activity.Mod$Steps.Interval.Avg,Data.Activity.Mod$steps)
```

A new data frame "Data.Activity.Mod" is created where NAs are replaced with average number of steps for the similar interval in the entire data.

```{r}
Steps.Day.Summary.Mod=ddply(Data.Activity.Mod,"date",summarise,Total.Steps.InDay.Mod=sum(steps))
#print(Steps.Day.Summary.Mod)
B=as.table(summary(Steps.Day.Summary$Total.Steps))
binwidth.mod=(B[6]-B[1])/20#(Max-Min)/200
qplot(Total.Steps.InDay.Mod,data = Steps.Day.Summary.Mod,geom = "histogram",binwidth=binwidth.mod,main="Histogram of Steps/Day after replacing NAs",xlab = "Steps",ylab = "Count (Days)",fill=I("blue"), col=I("red"))
```

To calcualte mean and median of new data frame where NAs are replaced

```{r}
Mean.Steps.InDay.Mod=mean(Steps.Day.Summary.Mod$Total.Steps.InDay.Mod)
Median.Steps.InDay.Mod=median(Steps.Day.Summary.Mod$Total.Steps.InDay.Mod)
```
After replacing NAs the average (mean) number of steps per day is `r Mean.Steps.InDay.Mod` (`r Mean.Steps.InDay` when NAs are discarded) and Median steps per day is `r Median.Steps.InDay.Mod` (`r Median.Steps.InDay` when NAs are discarded). We see that there is no much change in mean value where is the median changed by approximately 4 steps.

## Weekday and Weekend statistics
```{r}
Data.Activity.Mod$date=as.Date(Data.Activity.Mod$date)
Data.Activity.Mod$Day.Name=weekdays(Data.Activity.Mod$date)
Data.Activity.Mod$Week.Day.Or.End=ifelse(Data.Activity.Mod$Day.Name %in% c("Saturday","Sunday"),"Weekend","Weekday")
Steps.WeekDay.Summary=ddply(Data.Activity.Mod,.(interval,Week.Day.Or.End),summarise,Average.Steps=mean(steps))
Steps.WeekDay.Summary$interval=as.integer(as.character(Steps.WeekDay.Summary$interval))
g1<-ggplot(data = Steps.WeekDay.Summary,aes(x=interval,y=Average.Steps))+geom_line()
g1+facet_grid(Week.Day.Or.End ~ .)+labs(title="Avergae weekday vs weekend activity per interval",x="Interval",y="Average number of steps")
```